---
title: "GOV51 PSET2"
author: "\\small https://github.com/Steffny2004/gov51-ps2"
subtitle: "\\large Steffny Meegoda"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
	echo = TRUE,
	eval = TRUE,
	message = FALSE,
	warning = FALSE
)
```


# 1 Part 1: The Original Study (20 points)

## 1.1 Exploring the Experiment

### 1.1.1 Load packages and data

```{r}
library(tidyverse)
gay <- read_csv("data/raw/gay.csv")
```

### Q1.1 (3 points): How many total observations are in `gay.csv`? How many unique values does `study` take? List all the `treatment` conditions in the treatment variable.

```{r}
n <- nrow(gay)
unique_study_count <- length(unique(gay$study))
treatment_conditions <- unique(gay$treatment)

cat("There are", n, "total observations in gay.csv.\n\n")
cat("The study variables takes on", unique_study_count, "unique values.\n\n")
cat("The treatment conditions in the treatment variable are:\n", 
    paste(treatment_conditions, collapse = "\n "))
```

### Q1.2 (3 points): Filter to Study 1, Wave 1 only. Calculate the mean of ssm (same-sex marriage attitude, scored 1–5) for each treatment group using group_by() and summarize(). Which treatment group shows the highest average support?

```{r}
# Filter to Study 1, Wave 1
gay_s1w1 <- gay %>%
  filter(study == 1 & wave == 1)

# Calculate the mean ssm
gay_s1w1_mean_ssm <- gay_s1w1 %>%
  group_by(treatment) %>%
  summarize(mean_ssm = mean(ssm)) %>%
  arrange(desc(mean_ssm))
gay_s1w1_mean_ssm
```

*The Recycling Script by Gay Canvasser group showed the highest average support of 3.13.*

### Q1.3 (3 points): Still using Study 1, Wave 1, calculate the standard deviation of ssm within each treatment group. Are the standard deviations roughly similar across groups? Why is this important in a randomized experiment? (2–3 sentences)

```{r}
# Calculate the sd ssm
gay_s1w1_sd_ssm <- gay_s1w1 %>%
  group_by(treatment) %>%
  summarize(sd_ssm = sd(ssm)) %>%
  arrange(desc(sd_ssm))
gay_s1w1_sd_ssm
```

*We can see that the standard deviations for the ssm's across the treatment groups looks very similar, with all of them being close to 1.68. It's important that the standard deviations are roughly similar across groups because then we can more confidently say that the observed differences between treatment groups are caused by the treatment itself rather than unequal variability between groups.*

### Q1.4 (3 points): Create a bar plot of the mean ssm by treatment group (Study 1, Wave 1) using ggplot() with geom_col(). Hint: First create a summary data frame with the means, then plot it.

```{r}
library(scales)

ggplot(data = gay_s1w1_mean_ssm, aes(x = treatment, y = mean_ssm, fill = treatment)) + 
  geom_col() + 
  labs(
    title = "Mean Same-Sex Marriage Attitudes by Canvassing Group",
    subtitle = "Study 1, Wave 1", 
    x = "Canvassing Groups",
    y = "Mean SSM") +
  theme_minimal() + 
  theme(legend.position="none") + 
  geom_text(aes(label = round(mean_ssm, 2), vjust = -0.5), color = "black") + 
  scale_x_discrete(labels = label_wrap(width = 20)) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 3.5)
```

### 1.2 Checking Balance and Comparing Outcomes

### Q1.5 (4 points): A well-run randomized experiment should produce groups that look similar at baseline. Filter gay.csv to Study 1, Wave 1. Compare the “Same-Sex Marriage Script by Gay Canvasser” group and the “No Contact” group:

-   Calculate the mean and SD of ssm for each group
-   Compute the standardized difference in means: $\frac{{\bar{x}}_{\text{gay canvasser}} - {\bar{x}}_{\text{no contact}}}{\sqrt{s^2_{\text{gay canvasser}} + s^2_{\text{no contact}}} / 2}$

A standardized difference below 0.25 in absolute value suggests good balance. Do these groups look similar at baseline? (2–3 sentences)

```{r}
gay_compare_s1w1 <- gay_s1w1 %>%
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser", "No Contact")) %>%
  group_by(treatment) %>%
  summarize(
    mean_ssm = mean(ssm), 
    sd_ssm = sd(ssm))
gay_compare_s1w1

standardized_mean_diff_s1w1 <- (gay_compare_s1w1$mean_ssm[2] - gay_compare_s1w1$mean_ssm[1]) / 
  sqrt((((gay_compare_s1w1$sd_ssm[2])^2 + (gay_compare_s1w1$sd_ssm[1])^2) / 2))
```

*The standardized difference in means between the “Same-Sex Marriage Script by Gay Canvasser” group and the “No Contact” groups is `r standardized_mean_diff_s1w1', which has an absolute value below 0.25 and suggests good balance. Looking at our summary table, it agrees with this metric intuitively given that at baseline, these two groups seem to be very similar in terms of balance as their mean ssm's during Wave 1 are so close to one another even before standardization (0.017569 unstandardized difference), and definitely overlap within two standard errors of each other*

### Q1.6 (4 points): Now look at the outcomes. Filter to Study 1, Wave 2 only. Calculate the mean ssm for the gay canvasser group and the No Contact group. What is the difference in means (gay canvasser minus no contact)?

Interpret: since the groups were balanced at baseline, what does a difference at Wave 2 suggest? (2–3 sentences)

```{r}
gay_s1w2 <- gay %>%
  filter(study == 1 & wave == 2)
  
gay_compare_s1w2 <- gay_s1w2 %>%
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser", "No Contact")) %>%
  group_by(treatment) %>%
  summarize(
    mean_ssm = mean(ssm), 
    sd_ssm = sd(ssm))
gay_compare_s1w2

standardized_mean_diff_s1w2 <- (gay_compare_s1w2$mean_ssm[2] - 
                                  gay_compare_s1w2$mean_ssm[1]) / 
  sqrt((((gay_compare_s1w2$sd_ssm[2])^2 + (gay_compare_s1w2$sd_ssm[1])^2) / 2))
```

*Given that we considered the groups balanced at baseline, the standardized difference now being `r standardized_mean_diff_s1w2` is very comparable to baseline given that it is still below 0.25 in absolute value. Although the standardized difference in means is 0.05 greater than it was at baseline, this change still seems minuscule and suggests a balance between the groups.*
