---
title: "GOV51 PSET2"
author: "\\small https://github.com/Steffny2004/gov51-ps2"
subtitle: "\\large Steffny Meegoda"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
	echo = TRUE,
	eval = TRUE,
	message = FALSE,
	warning = FALSE
)
```

# 1 Part 1: The Original Study (20 points)

## 1.1 Exploring the Experiment

### 1.1.1 Load packages and data

```{r}
library(tidyverse)
gay <- read_csv("data/raw/gay.csv")
```

### Q1.1 (3 points): How many total observations are in `gay.csv`? How many unique values does `study` take? List all the `treatment` conditions in the treatment variable.

```{r}
n <- nrow(gay)
unique_study_count <- length(unique(gay$study))
treatment_conditions <- unique(gay$treatment)

# Display results
cat("Total observations:", n, 
    "\n\nNumber of unique studies:", unique_study_count,
    "\n\nThe treatment conditions in the treatment variable are:\n-", 
    paste(treatment_conditions, collapse = "\n- "))
```

### Q1.2 (3 points): Filter to Study 1, Wave 1 only. Calculate the mean of ssm for each treatment group. Which treatment group shows the highest average support?

```{r}
# Filter to Study 1, Wave 1
gay_s1w1 <- gay %>%
  filter(study == 1 & wave == 1)

# Calculate the mean ssm
gay_s1w1_mean_ssm <- gay_s1w1 %>%
  group_by(treatment) %>%
  summarize(
    mean_ssm = mean(ssm), 
    n = n()) %>%
  arrange(desc(mean_ssm))
gay_s1w1_mean_ssm
```

*The Recycling Script by Gay Canvasser group showed the highest average support of 3.13.*

\clearpage

### Q1.3 (3 points): Still using Study 1, Wave 1, calculate the standard deviation of ssm within each treatment group. Are the standard deviations roughly similar across groups? Why is this important in a randomized experiment? (2–3 sentences)

```{r}
# Calculate the sd ssm
gay_s1w1_sd_ssm <- gay_s1w1 %>%
  group_by(treatment) %>%
  summarize(sd_ssm = sd(ssm)) %>%
  arrange(desc(sd_ssm))
gay_s1w1_sd_ssm
```

*We can see that the standard deviations for the ssm's across the treatment groups looks very similar, with all of them being close to 1.68. It's important that the standard deviations are roughly similar across groups because then we can more confidently say that the observed differences between treatment groups are caused by the treatment itself rather than unequal variability between groups.*

\clearpage

### Q1.4 (3 points): Create a bar plot of mean ssm by treatment group (Study 1, Wave 1).

```{r}
library(scales)

ggplot(data = gay_s1w1_mean_ssm, aes(x = treatment, y = mean_ssm, fill = treatment)) + 
  geom_col() + 
  labs(
    title = "Mean Same-Sex Marriage Attitudes by Canvassing Group",
    subtitle = "Study 1, Wave 1", 
    x = "Canvassing Groups",
    y = "Mean SSM") +
  theme_minimal() + 
  theme(legend.position="none") + 
  geom_text(aes(label = round(mean_ssm, 2), vjust = -0.5), color = "black") + 
  scale_x_discrete(labels = label_wrap(width = 20)) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 3.5)
```

\clearpage

## 1.2 Checking Balance and Comparing Outcomes

### Q1.5 (4 points): Filter gay.csv to Study 1, Wave 1. Compare the “Same-Sex Marriage Script by Gay Canvasser” group and the “No Contact” group:

-   Calculate the mean and SD of ssm for each group
-   Compute the standardized difference in means: $\frac{{\bar{x}}_{\text{gay canvasser}} - {\bar{x}}_{\text{no contact}}}{\sqrt{s^2_{\text{gay canvasser}} + s^2_{\text{no contact}}} / 2}$

A standardized difference below 0.25 in absolute value suggests good balance. Do these groups look similar at baseline? (2–3 sentences)

```{r}
gay_s1w1_compare <- gay_s1w1 %>%
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser", "No Contact")) %>%
  group_by(treatment) %>%
  summarize(
    mean_ssm = mean(ssm), 
    sd_ssm = sd(ssm))
gay_s1w1_compare

standardized_mean_diff_s1w1 <- (gay_s1w1_compare$mean_ssm[2] - 
                                  gay_s1w1_compare$mean_ssm[1]) / 
  sqrt((((gay_s1w1_compare$sd_ssm[2])^2 + (gay_s1w1_compare$sd_ssm[1])^2) / 2))

cat("Standardized difference in means:", standardized_mean_diff_s1w1)
```

*The standardized difference in means between the “Same-Sex Marriage Script by Gay Canvasser” group and the “No Contact” groups is `r standardized_mean_diff_s1w1`, which has an absolute value below 0.25 and suggests good balance. Looking at our summary table, it agrees with this metric intuitively given that at baseline, these two groups seem to be very similar in terms of balance as their mean ssm's during Wave 1 are so close to one another even before standardization (0.017569 unstandardized difference), and definitely overlap within two standard errors of each other.*

\clearpage

### Q1.6 (4 points): Now look at the outcomes. Filter to Study 1, Wave 2 only. Calculate the mean ssm for each group. What is the difference in means (gay canvasser - no contact)? 

Since the groups were balanced at baseline, what does a difference at Wave 2 suggest? (2–3 sentences)

```{r}
gay_s1w2 <- gay %>%
  filter(study == 1 & wave == 2)
  
gay_s1w2_compare <- gay_s1w2 %>%
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser", "No Contact")) %>%
  group_by(treatment) %>%
  summarize(
    mean_ssm = mean(ssm), 
    sd_ssm = sd(ssm))
gay_s1w2_compare

standardized_mean_diff_s1w2 <- (gay_s1w2_compare$mean_ssm[2] - 
                                  gay_s1w2_compare$mean_ssm[1]) / 
  sqrt((((gay_s1w2_compare$sd_ssm[2])^2 + (gay_s1w2_compare$sd_ssm[1])^2) / 2))

cat("Standardized difference in means:", standardized_mean_diff_s1w2)
```

*Given that we considered the groups balanced at baseline, the standardized difference now being `r standardized_mean_diff_s1w2` is very comparable to baseline as it is still below 0.25 in absolute value. Although the standardized difference in means is 0.05 greater than it was at baseline, this change still seems minuscule and suggests a balance between the groups.*

\clearpage

# 2 Part 2: The Forensic Discovery (30 points)

## 2.1 The Baseline Distribution

### Q2.1 (3 points): Load `gayreshaped.csv` and `ccap2012.csv.` Report the number of observations and column names for each dataset.

```{r}
# Load data
gayr <- read_csv("data/raw/gayreshaped.csv")
ccap <- read_csv("data/raw/ccap2012.csv")

# Count observations and find colnames
n_gayr <- nrow(gayr)
colnames_gayr <- colnames(gayr)
n_ccap <- nrow(ccap)
colnames_ccap <- colnames(ccap)

# Display results
cat("gayreshaped.csv:\n", n_gayr, "total observations and the column names are\n-", 
    paste(colnames_gayr, collapse = "\n- "))

cat("ccap2012.csv.:\n", n_ccap, "total observations and the column names are\n-", 
    paste(colnames_ccap, collapse = "\n- "))
```

### Q2.2 (5 points): Extract the baseline feeling thermometer (`therm1`) for Study 1 from `gayreshaped.csv.` Calculate its mean, median, and standard deviation. Then do the same for gaytherm from `ccap2012.csv` (remove NAs). Report both sets of statistics side by side. What do you notice? (1–2 sentences)

```{r}
gayr_s1 <- gayr %>%
  filter(study == 1) %>%
  drop_na(therm1)

gayr_s1_summary <- gayr_s1 %>%
  summarize(
    mean= mean(therm1), 
    median = median(therm1), 
    sd = sd(therm1))

ccap_baseline_summary <- ccap %>%
  drop_na(gaytherm) %>%
  summarize(
    mean = mean(gaytherm), 
    median = median(gaytherm), 
    sd = sd(gaytherm))

# Report side-by-side
combined_data_summary <- tibble(
  stat = c("Mean", "Median", "SD"),
  gayr = c(
    gayr_s1_summary$mean,
    gayr_s1_summary$median,
    gayr_s1_summary$sd
  ),
  ccap = c(
    ccap_baseline_summary$mean,
    ccap_baseline_summary$median,
    ccap_baseline_summary$sd
  )
)

# Display results
combined_data_summary
```

*Looking at the test statistics of both datasets side by side, it seems that the mean, median, and sd between the two are extremely similar in terms of thermometer scoring and have lots of overlap. It would appear that their samples are pretty balanced and similar to one another.*

### Q2.3 (5 points): Create side-by-side histograms comparing the LaCour Study 1 baseline (therm1) and the CCAP data (gaytherm).

Do these look like they came from different populations? (2–3 sentences)

```{r}
# Combine datasets
combined_data <- bind_rows(
  gayr_s1 %>%
    transmute(therm = therm1, source = "LaCour Study 1"),
  ccap %>%
    drop_na(gaytherm) %>%
    transmute(therm = gaytherm, source = "CCAP"))

# Plot side-by-side histograms
ggplot(combined_data, aes(x = therm)) +
  geom_histogram(breaks = seq(0, 100, by = 5), fill = "lightblue", color = "skyblue") +
  facet_wrap(~ source) +
  labs(
    title = "Baseline Feeling Thermometer: LaCour Study 1 vs CCAP",
    x = "Feeling Thermometer",
    y = "Count") +
  theme_minimal()
```

*The distributions of both datasets look very similar, with the main difference being in the number of observations. Otherwise, the heaping towards the middle value at 50 and the heaping towards the more extreme ends of 0 and 100 track across both distributions, making them look like they came from very similar populations.*

\clearpage

## 2.2 Re-test Reliability

### Q2.4 (6 points): For both Study 1 and Study 2 control groups, calculate:

(a) The correlation between `therm1` and `therm2` (use `cor()` with use = "`complete.obs`")
(b) The standard deviation of the within-person change (`therm2` - `therm1`)

Present your results side by side. A re-test correlation of 0.95–0.97 is typical for real feeling thermometer data. Which study shows a more plausible pattern, and why? (3–4 sentences)

```{r}
gayr_nc_s1_summary <- gayr %>%
  drop_na(therm1, therm2) %>%
  filter(study == 1 & treatment == "No Contact") %>%
  summarize(
    cor = cor(therm1, therm2, use =  "complete.obs"), 
    sd = sd(therm2 - therm1))
  
gayr_nc_s2_summary <- gayr %>%
  drop_na(therm1, therm2) %>%
  filter(study == 2 & treatment == "No Contact") %>%
  summarize(
    cor = cor(therm1, therm2, use =  "complete.obs"), 
    sd = sd(therm2 - therm1))

# Report side-by-side
combined_s1_s2_summary <- tibble(
  stat = c("Correlation", "SD of Within-Person Change"),
  gayr_control_s1 = c(gayr_nc_s1_summary$cor, gayr_nc_s1_summary$sd),
  gayr_control_s2 = c(gayr_nc_s2_summary$cor, gayr_nc_s2_summary$sd)
  )
combined_s1_s2_summary
```

*Study 2 shows a more plausible pattern because it's correlation isn't as high (0.972) as it is for Study 1 (0.998), which in Study 1 seems to be too good to be true. Also, it's standard deviation of the within-person change between `therm1` and `therm2` is larger than it was initially.*

\clearpage

### Q2.5 (5 points): Create two scatter plots side by side:

(a) Study 1 control group: `therm1` (x-axis) vs `therm2` (y-axis), with a 45-degree line
(b) Study 2 control group: `therm1` (x-axis) vs `therm2` (y-axis), with a 45-degree line

Describe the differences. Which looks more like real longitudinal data, and why? (3–4 sentences)

```{r}
gayr_nc_s1 <- gayr %>%
  drop_na(therm1, therm2) %>%
  filter(study == 1 & treatment == "No Contact")

gayr_nc_s2 <- gayr %>%
  drop_na(therm1, therm2) %>%
  filter(study == 2 & treatment == "No Contact")
  
# Combine datasets
combined_nc_s1s2 <- bind_rows(
  gayr_nc_s1 %>%
    transmute(therm1 = therm1, therm2 = therm2, source = "Study 1"),
  gayr_nc_s2 %>%
    transmute(therm1 = therm1, therm2 = therm2, source = "Study 2"))

# Plot side-by-side histograms
ggplot(combined_nc_s1s2, aes(x = therm1, y = therm2)) +
  geom_point(color = "lightblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "blue", linetype = "solid") +
  facet_wrap(~ source) +
  labs(
    title = "Correlation Between `therm1` and `therm2`",
    x = "Feeling Thermometer",
    y = "Count") +
  theme_minimal()
```

*When looking at the correlation between `therm1` and `therm2`, Study 1 shows nearly perfect positive linear correlation as the data falls barely above and below our $45^{\circ}$ line. In contrast, Study 2 shows positive linear correlation that is not as strong, as depicted by the jitter in the data above and below the $45^{\circ}$, and with the data being more spread out or sparse in some regions of the Feeling Thermometer. It is clear that Study 2's scatterplot looks like real longitudinal data in comparison to Study 1 because of the natural noise that seems to exist in Study 2, but in addition to the neatness along the line, the heaping in the middle for Study 1 on the scatterplot looks unnaturally piled along the center in comparison to Study 2.*

\clearpage

## 2.3 Putting It All Together

### Q2.6 (6 points): You’ve now found three irregularities:

1.  The baseline distributions match an existing national survey
2.  The re-test correlations are implausibly high (Study 1)
3.  The within-person changes are implausibly small (Study 1)

In 4–5 sentences, explain why these three findings together suggest the data were fabricated rather than collected in a real experiment. Use specific numbers from your analysis. Why would a fabricator produce data with these properties?

*Taken together, these three patterns strongly suggest the data were fabricated rather than collected from real respondents. First, the baseline distribution in Study 1 closely matches the national CCAP survey, with nearly identical summary statistics, which would be unlikely if an independent sample was taken for LaCour's study, so it is more consistent with copying the already existing dataset. Second, the re-test correlation in Study 1 is almost impossibly high at 0.998, which flags as weird because real attitude measures over time typically show much lower stability due to measurement error and natural variation. Third, the within-person changes are extremely small, with differences clustered near zero and a very small standard deviation, which is unrealistic in real panel data where individuals normally vary more. A fabricator might generate data this way to make the study look clean and convincing—highly stable attitudes with minimal noise, while also using a known national distribution to ensure the baseline appears realistic.*

\clearpage

# 3 Part 3: What Real Data Looks Like (25 points)

## 3.1 Comparing Baselines

### Q3.1 (3 points): Filter gayreshaped.csv to Study 2. How many observations? Calculate the mean and SD of therm1 (baseline) for the treatment group and the control group separately.

```{r}
gayr_s2 <- gayr %>%
  filter(study == 2)

gayr_s2_compare_baseline <- gayr_s2 %>%
  drop_na(therm1) %>%
  group_by(treatment) %>%
  summarize(
    mean_therm1 = mean(therm1), 
    sd_therm1 = sd(therm1), 
    n = n())
gayr_s2_compare_baseline
```

### Q3.2 (3 points): Create side-by-side histograms of Study 1 baseline (`therm1`) and Study 2 baseline (`therm1`). How do they compare visually? (2–3 sentences)

```{r}
# Combine datasets
combined_gayr_s1s2 <- bind_rows(
  gayr_s1 %>%
    drop_na(therm1) %>%
    transmute(therm = therm1, source = "Study 1"),
  gayr_s2 %>%
    drop_na(therm1) %>%
    transmute(therm = therm1, source = "Study 2"))

# Plot side-by-side histograms
ggplot(combined_gayr_s1s2, aes(x = therm)) +
  geom_histogram(breaks = seq(0, 100, by = 5), fill = "lightblue", color = "skyblue") +
  facet_wrap(~ source) +
  labs(
    title = "Baseline Feeling Thermometer: Study 1 vs Study 2", 
    x = "Feeling Thermometer",
    y = "Count") +
  theme_minimal()
```

*Visually, the distributions of both datasets look very similar to how the CCAP and LaCour Study 1 comparison looked, but now with Study 1 having more observations than study 2. The similarities in the heaping towards the middle value at 50 and the heaping towards the more extreme ends of 0 and 100 track across both distributions, again making them look like they came from very similar populations.*

## 3.2 Checking Balance and Comparing Outcomes in Study 2

### Q3.3 (3 points): Filter gayreshaped.csv to Study 2, keeping only the “Same-Sex Marriage Script by Gay Canvasser” and “No Contact” groups. Check balance at baseline: calculate the mean and SD of therm1 for each group and the standardized difference in means (same formula as Q1.5). Do the groups look balanced? (1–2 sentences)

```{r}
gayr_s2_compare_baseline <- gayr_s2 %>%
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser", "No Contact")) %>%
  drop_na(therm1) %>%
  group_by(treatment) %>%
  summarize(
    mean_therm1 = mean(therm1), 
    sd_therm1 = sd(therm1))
gayr_s2_compare_baseline

standardized_mean_diff_gayr_s2 <- (gayr_s2_compare_baseline$mean_therm1[2] - 
                                     gayr_s2_compare_baseline$mean_therm1[1]) / 
  sqrt(((gayr_s2_compare_baseline$sd_therm1[2])^2 + 
          (gayr_s2_compare_baseline$sd_therm1[1])^2) / 2)

cat("Standardized difference in means:", standardized_mean_diff_gayr_s2)
```

*The standardized difference in means between the “Same-Sex Marriage Script by Gay Canvasser” group and the “No Contact” groups for Study 2 now is `r standardized_mean_diff_gayr_s2`, which has an absolute value very below 0.25 and suggests good balance. Looking at our summary table, it also agrees with this metric intuitively given that at baseline, these two groups seem to be very similar in terms of balance as their mean ssm's during Wave 1 are so close to one another even before standardization (0.28077 unstandardized difference), and definitely overlap within two standard errors of each other.*

### Q3.4 (3 points): Now compare outcomes. Calculate the mean `therm2` (post-conversation) for each group. What is the difference in post-period means (gay canvasser minus no contact)?

```{r}
gayr_s2_compare_outcome <- gayr_s2 %>%
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser", "No Contact")) %>%
  drop_na(therm2) %>%
  group_by(treatment) %>%
  summarize(
    mean_therm2 = mean(therm2), 
    sd_therm2 = sd(therm2))
gayr_s2_compare_outcome

mean_diff_gayr_s2 <- gayr_s2_compare_outcome$mean_therm2[2] - 
  gayr_s2_compare_outcome$mean_therm2[1]

cat("The difference in post-period means is", mean_diff_gayr_s2)
```

\clearpage

## 3.3 Uncertainty Around the Difference

### Q3.5 (4 points): Calculate the standard error of the difference in post-period means from Q3.4. Use the formula:

$$SE = \sqrt{\frac{s^2_{\text{gay canvasser}}}{n_{\text{gay canvasser}}} + \frac{s^2_{\text{no contact}}}{n_{\text{no contact}}}}$$ 
**where $s^2$ is the variance of `therm2` and $n$ is the sample size for each group. Then construct a 95% confidence interval: $(\bar{x_1} - \bar{x_2}) \pm 1.96 \times SE$. Does the CI include zero? What does this tell you about whether the groups differ after the conversations?**

```{r}
se_parameters <- gayr_s2 %>%
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser", "No Contact")) %>%
  drop_na(therm2) %>%
  group_by(treatment) %>%
  summarize(
    var = var(therm2),
    n = n())

se <- sqrt((se_parameters$var[2]) / (se_parameters$n[2]) + 
             (se_parameters$var[1]) / (se_parameters$n[1]))

ci_upper <- mean_diff_gayr_s2 + 1.96 * se
ci_lower <- mean_diff_gayr_s2 - 1.96 * se

# Display results
cat("SE =", se, 
    "\n95% CI:", ci_lower, ",", ci_upper)
```

*The standard error of the difference in post-period means from Q3.4 is `r se`, and the 95% CI is [`r ci_upper`, `r ci_lower`]. As we see, the 95% CI does not include 0, which suggests that the difference between the two groups is non-zero and that there is indeed some treatment effect difference after the conversations where there is a higher thermometer score for the gay canvassers post-conversation compared to the no contact groups.*

\clearpage

## 3.4 Tracking Attitudes Over Time

### Q3.6 (4 points): For the Study 2 control group, calculate the correlation between therm1 and each subsequent wave: therm2, therm3, and therm4. Use cor() with use = "complete.obs". Do the correlations decay over time (i.e., do more distant waves correlate less with baseline)? What does this pattern tell you about the stability of real survey measurements? (2–3 sentences)

```{r}
gayr_nc_s2_compare_waves <- gayr_nc_s2 %>%
  summarize(
    cor_therm12 = cor(therm1 , therm2, use = "complete.obs"),
    cor_therm13 = cor(therm1 , therm3, use = "complete.obs"),
    cor_therm14 = cor(therm1 , therm4, use = "complete.obs"))
gayr_nc_s2_compare_waves
```

*The correlations between therm 1 and each subsequent wave appears to oscillate, with there being a high correlation of `r gayr_nc_s2_compare_waves$cor_therm12` after the second wave, then this correlation decreases to `r gayr_nc_s2_compare_waves$cor_therm13` after the third wave, and then increases slightly back to `r gayr_nc_s2_compare_waves$cor_therm14` after the fourth wave. So it seems that the more distant waves correlate less with baseline, and indicate that there is a decent amount of variability within real survey data.*

\clearpage

### Q3.7 (5 points): You’ve now analyzed fabricated data (Study 1), caught the fraud, and examined more realistic data (Study 2). In 5–6 sentences, discuss:

-   What makes Study 2’s data different from Study 1’s data? Be specific about which statistics differ and by how much.

-   Why is it important that real treatment effects are smaller and messier than fabricated ones?

-   What role did simple measurement tools (mean, SD, histograms, correlations) play in detecting the fraud?

*Study 2’s data looks different and more realistic than Study 1’s data because the treatment effects are smaller and the data is naturally noisier. In Study 2, the baseline means of `therm1` for the treatment and control groups are very similar (as there is only a difference of 0.28077 points with a standardized difference well below 0.25), which indicates that there was indeed proper randomization, whereas Study 1 showed unusually neat and precise patterns. The post-treatment difference in Study 2 is also smaller and less exaggerated than the perfectly consistent effects we saw in Study 1, which makes sense since the treatment effect would be smaller and messier due to natural human variation human in attitudes, measurement errors in surveys, and the natural fluctuation of people’s responses over time. This variability is clear in Study 2 as seen by the large standard deviations and correlations that change across waves, in comparison to Study 1 where the correlation remained rather high throughout. Simple tools like means, standard deviations, histograms, and correlations were important for detecting the fraud because they revealed patterns in Study 1 that were inconsistent with real data by being too smooth and consistent.*

\clearpage

# 4 Part 4: Formalizing the Evidence with Regression (25 points)

## 4.1 Regression as a Fraud Detector

### Q4.1 (5 points): For the Study 1 control group, fit a linear regression predicting therm2 from `therm1`:

$$therm2_i = \beta_0 + \beta_1 \times therm1_i + \epsilon_i$$

Store the result as `fit_s1` and display the summary. Report the slope ($\hat{\beta_1}$), the R-squared, and the residual standard error.

```{r}
fit_s1 <- lm(therm2 ~ therm1, data = gayr_nc_s1)
summary(fit_s1)
```

```{r}
# Display results
cat("B1_hat:", summary(fit_s1)$coefficients[2], 
    "\nR-squared:",summary(fit_s1)$r.squared, 
    "\nRSE:", summary(fit_s1)$sigma)
```


### Q4.2 (5 points): Repeat Q4.1 for the Study 2 **control group**. Store as `fit_s2.` Report the same three quantities: slope, R-squared, and residual standard error. Compare the two models. In fabricated data, we’d expect $\hat{\beta_1} \approx 1$ and $R^2 \approx 1$ because the “follow- up” data is just a copy of the baseline with tiny noise added. In real data, we’d expect regression to the mean: $\hat{\beta_1} < 1$ and lower $R^2$. Which pattern do you see? (3–4 sentences)

```{r}
fit_s2 <- lm(therm2 ~ therm1, data = gayr_nc_s2)
summary(fit_s2)
```

```{r}
# Display results
cat("B1_hat:", summary(fit_s2)$coefficients[2], 
    "\nR-squared:",summary(fit_s2)$r.squared, 
    "\nRSE:", summary(fit_s2)$sigma)
```

*For Study 2, the regression of `therm2` on `therm1` in the control group gives us a slope of 0.969, an $R_2$ of 0.948, and a residual standard error of 6.55. Compared to Study 1, where we had a slope of 0.992$, an $R_2$ of 0.995, and a residual standard error of 1.98), the relationship between the baseline and subsequent wave attitudes is weaker and noisier. In Study 2, the slope $\hat{\beta_1}$ being less than 1 shows regression to the mean and the lower $R^2$ in Study 2 shows that baseline attitudes explain less of the variation in the subsequent wave, which is consistent with real data. Overall, Study 2 displays the pattern expected in real data, whereas Study 1’s near perfect fit with both $\hat{\beta_1}$ and $R^2$ are approximately 1 is suspicious and consistent with fabricated data.*

## 4.2 Comparing Groups with Regression

### Q4.3 (5 points): Now use regression to compare the two groups in Study 2. Create a binary variable treated that equals 1 for the “Same-Sex Marriage Script by Gay Canvasser” group and 0 for the “No Contact” group. Then fit:

$$therm2_i = \beta_0 + \beta_1 \times treated_i + \epsilon_i$$

Report $\hat{\beta_1}$, its standard error, t-statistic, and p-value. Compare $\hat{\beta_1}$ to the difference in means you calculated in Q3.4 — what do you notice? What does $\hat{\beta_0}$ estimate? Is the difference between groups statistically significant at $\alpha$ = 0.05?

```{r}
gayr_s2_binary <- gayr_s2 %>%
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser", "No Contact")) %>%
  mutate(treated = ifelse(treatment == "Same-Sex Marriage Script by Gay Canvasser", 1, 0))

fit_s2_binary <- lm(therm2 ~ treated, data = gayr_s2_binary)
summary(fit_s2_binary)
```

```{r}
# Display results
cat("Beta1_hat:", summary(fit_s2_binary)$coefficients["treated", "Estimate"],
    "\nSE:", summary(fit_s2_binary)$coefficients["treated", "Std. Error"],
    "\nt-stat:", summary(fit_s2_binary)$coefficients["treated", "t value"],
    "\np-value:", summary(fit_s2_binary)$coefficients["treated", "Pr(>|t|)"])
```

*This regression gives us the estimates $\hat{\beta_1}$ = 3.568, $SE$ = 1.243, $t$ = 2.872, and a p-value of $0.0041$. The coefficient estimate, $\hat{\beta_1}$, represents the difference in mean post-treatment thermometer scores between the treated group and the control group, and it matches the difference in means calculated in Q3.4 because a regression with a binary treatment variable is mathematically equivalent to a difference-in-means comparison. The intercept, $\hat{\beta_0}$ estimates the average `therm2` score for the control (“No Contact”) group, since this is the group where `treated` = 0. Because the p-value is less than 0.05, the difference between the treated and control groups is statistically significant at the 5% level. This suggests that the canvassing treatment had a measurable positive effect on attitudes toward same-sex marriage in Study 2.*

## 4.3 Controlling for Baseline Attitudes

### Q4.4 (5 points): Add therm1 as a control variable:

$$therm2_i = \beta_0 + \beta_1 \times treated + \beta_2 \times therm1_i +  \epsilon_i$$ Report the coeﬀicient on `treated`, its standard error, t-statistic, and p-value. Compare this to your results from Q4.3:

(a) Did the estimate of $\hat{\beta_1}$ change much?
(b) Did the standard error change? In which direction?
(c) Why does controlling for baseline attitudes improve precision? (2–3 sentences)

```{r}
fit_s2_binary_therm1 <- lm(therm2 ~ treated + therm1, data = gayr_s2_binary)
summary(fit_s2_binary_therm1)
```

```{r}
# Display results
cat("Beta1_hat:", summary(fit_s2_binary_therm1)$coefficients["treated", "Estimate"],
    "\nSE:", summary(fit_s2_binary_therm1)$coefficients["treated", "Std. Error"],
    "\nt-stat:", summary(fit_s2_binary_therm1)$coefficients["treated", "t value"],
    "\np-value:", summary(fit_s2_binary_therm1)$coefficients["treated", "Pr(>|t|)"])
```

*Compared to Q4.3, the estimate of $\hat{\beta_1}$ decreased from 3.568 to 2.459, but we still see a positive treatment effect that is considered significant. The standard error decreased substantially from 1.242 to 0.341, which is a 73% reduction of standard error that makes the estimate much more precise and increases the t-statistic. Controlling for baseline attitudes improves precision because `therm1` explains a large portion of the variation in `therm2`, which reduces unexplained noise in the model. By accounting for individuals’ initial attitudes, the regression is able to capture the treatment effect better and gives us a more precise estimate.*

### Q4.5 (5 points): Compare the R-squared values from Q4.3 and Q4.4. Why does adding therm1 increase the R-squared so dramatically? What does this tell you about how much of the variation in follow-up attitudes is explained by where people started versus the treatment?

*The R-squared increases from 0.0039 in Q4.3 to 0.925 in Q4.4 after adding therm1. This happens because baseline `therm1` scores strongly predict thermometer scores in subsequent waves, so most of the variation in `therm2` is therefore explained by where people started rather than by the treatment. The treatment still has an effect, but baseline attitudes explain more of the overall variation.*

Finally, use `predict()` to generate predicted values from your Q4.4 model for two hypothetical respondents:

-   Person A: `treated` = 1, `therm1` = 50 (moderate baseline)
-   Person B: `treated` = 0, `therm1` = 50 (same baseline, no treatment)

What is the predicted difference in their Wave 2 attitudes?

```{r}
# Hypothetical respondents
new_data <- data.frame(
  treated = c(1, 0),
  therm1 = c(50, 50))

# Predict
prediction <- predict(fit_s2_binary_therm1, newdata = new_data)
prediction

# Calculate predicted difference
pred_diff = prediction[1] - prediction[2]

cat("Predicted difference:", pred_diff)
```

\clearpage 

# 5 Summary

Nothing to write or submit for this problem. Submissions checklist completed.
