---
title: "GOV51 PSET2"
author: "\\small https://github.com/Steffny2004/gov51-ps2"
subtitle: "\\large Steffny Meegoda"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
	echo = TRUE,
	eval = TRUE,
	message = FALSE,
	warning = FALSE
)
```

# 1 Part 1: The Original Study (20 points)

## 1.1 Exploring the Experiment

### 1.1.1 Load packages and data

```{r}
library(tidyverse)
gay <- read_csv("data/raw/gay.csv")
```

### Q1.1 (3 points): How many total observations are in `gay.csv`? How many unique values does `study` take? List all the `treatment` conditions in the treatment variable.

```{r}
n <- nrow(gay)
unique_study_count <- length(unique(gay$study))
treatment_conditions <- unique(gay$treatment)

cat("There are", n, "total observations in gay.csv.\n\n")
cat("The study variables takes on", unique_study_count, "unique values.\n\n")
cat("The treatment conditions in the treatment variable are:\n-", 
    paste(treatment_conditions, collapse = "\n- "))
```

### Q1.2 (3 points): Filter to Study 1, Wave 1 only. Calculate the mean of ssm (same-sex marriage attitude, scored 1–5) for each treatment group using group_by() and summarize(). Which treatment group shows the highest average support?

```{r}
# Filter to Study 1, Wave 1
gay_s1w1 <- gay %>%
  filter(study == 1 & wave == 1)

# Calculate the mean ssm
gay_s1w1_mean_ssm <- gay_s1w1 %>%
  group_by(treatment) %>%
  summarize(mean_ssm = mean(ssm)) %>%
  arrange(desc(mean_ssm))
gay_s1w1_mean_ssm
```

*The Recycling Script by Gay Canvasser group showed the highest average support of 3.13.*

### Q1.3 (3 points): Still using Study 1, Wave 1, calculate the standard deviation of ssm within each treatment group. Are the standard deviations roughly similar across groups? Why is this important in a randomized experiment? (2–3 sentences)

```{r}
# Calculate the sd ssm
gay_s1w1_sd_ssm <- gay_s1w1 %>%
  group_by(treatment) %>%
  summarize(sd_ssm = sd(ssm)) %>%
  arrange(desc(sd_ssm))
gay_s1w1_sd_ssm
```

*We can see that the standard deviations for the ssm's across the treatment groups looks very similar, with all of them being close to 1.68. It's important that the standard deviations are roughly similar across groups because then we can more confidently say that the observed differences between treatment groups are caused by the treatment itself rather than unequal variability between groups.*

### Q1.4 (3 points): Create a bar plot of the mean ssm by treatment group (Study 1, Wave 1) using ggplot() with geom_col(). Hint: First create a summary data frame with the means, then plot it.

```{r}
library(scales)

ggplot(data = gay_s1w1_mean_ssm, aes(x = treatment, y = mean_ssm, fill = treatment)) + 
  geom_col() + 
  labs(
    title = "Mean Same-Sex Marriage Attitudes by Canvassing Group",
    subtitle = "Study 1, Wave 1", 
    x = "Canvassing Groups",
    y = "Mean SSM") +
  theme_minimal() + 
  theme(legend.position="none") + 
  geom_text(aes(label = round(mean_ssm, 2), vjust = -0.5), color = "black") + 
  scale_x_discrete(labels = label_wrap(width = 20)) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 3.5)
```

### 1.2 Checking Balance and Comparing Outcomes

### Q1.5 (4 points): A well-run randomized experiment should produce groups that look similar at baseline. Filter gay.csv to Study 1, Wave 1. Compare the “Same-Sex Marriage Script by Gay Canvasser” group and the “No Contact” group:

-   Calculate the mean and SD of ssm for each group
-   Compute the standardized difference in means: $\frac{{\bar{x}}_{\text{gay canvasser}} - {\bar{x}}_{\text{no contact}}}{\sqrt{s^2_{\text{gay canvasser}} + s^2_{\text{no contact}}} / 2}$

A standardized difference below 0.25 in absolute value suggests good balance. Do these groups look similar at baseline? (2–3 sentences)

```{r}
gay_compare_s1w1 <- gay_s1w1 %>%
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser", "No Contact")) %>%
  group_by(treatment) %>%
  summarize(
    mean_ssm = mean(ssm), 
    sd_ssm = sd(ssm))
gay_compare_s1w1

standardized_mean_diff_s1w1 <- (gay_compare_s1w1$mean_ssm[2] - gay_compare_s1w1$mean_ssm[1]) / 
  sqrt((((gay_compare_s1w1$sd_ssm[2])^2 + (gay_compare_s1w1$sd_ssm[1])^2) / 2))
```

*The standardized difference in means between the “Same-Sex Marriage Script by Gay Canvasser” group and the “No Contact” groups is \`r standardized_mean_diff_s1w1', which has an absolute value below 0.25 and suggests good balance. Looking at our summary table, it agrees with this metric intuitively given that at baseline, these two groups seem to be very similar in terms of balance as their mean ssm's during Wave 1 are so close to one another even before standardization (0.017569 unstandardized difference), and definitely overlap within two standard errors of each other*

### Q1.6 (4 points): Now look at the outcomes. Filter to Study 1, Wave 2 only. Calculate the mean ssm for the gay canvasser group and the No Contact group. What is the difference in means (gay canvasser minus no contact)?

Interpret: since the groups were balanced at baseline, what does a difference at Wave 2 suggest? (2–3 sentences)

```{r}
gay_s1w2 <- gay %>%
  filter(study == 1 & wave == 2)
  
gay_compare_s1w2 <- gay_s1w2 %>%
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser", "No Contact")) %>%
  group_by(treatment) %>%
  summarize(
    mean_ssm = mean(ssm), 
    sd_ssm = sd(ssm))
gay_compare_s1w2

standardized_mean_diff_s1w2 <- (gay_compare_s1w2$mean_ssm[2] - 
                                  gay_compare_s1w2$mean_ssm[1]) / 
  sqrt((((gay_compare_s1w2$sd_ssm[2])^2 + (gay_compare_s1w2$sd_ssm[1])^2) / 2))
```

*Given that we considered the groups balanced at baseline, the standardized difference now being `r standardized_mean_diff_s1w2` is very comparable to baseline given that it is still below 0.25 in absolute value. Although the standardized difference in means is 0.05 greater than it was at baseline, this change still seems minuscule and suggests a balance between the groups.*

\clearpage

# 2 Part 2: The Forensic Discovery (30 points)

## 2.1 The Baseline Distribution

### Q2.1 (3 points): Load `gayreshaped.csv` and `ccap2012.csv.` Report the number of observations and column names for each dataset. In `gayreshaped.csv`, the variables `therm1` through `therm4` represent feeling thermometer scores (0–100) toward gay men and lesbians, measured at each survey wave.

```{r}
# Load data
gay_r <- read_csv("data/raw/gayreshaped.csv")
ccap <- read_csv("data/raw/ccap2012.csv")

# Count observations and find colnames
n_gay_r <- nrow(gay_r)
colnames_gay_r <- colnames(gay_r)
n_ccap <- nrow(gay_r)
colnames_ccap <- colnames(ccap)

# Display results
cat("gayreshaped.csv:\n", n_gay_r, "total observations and the column names are\n-", paste(colnames_gay_r, collapse = "\n- "))

cat("\n\nccap2012.csv.:\n", n_ccap, "total observations and the column names are\n-", paste(colnames_ccap, collapse = "\n- "))

```

### Q2.2 (5 points): Extract the baseline feeling thermometer (`therm1`) for Study 1 from `gayreshaped.csv.` Calculate its mean, median, and standard deviation. Then do the same for gaytherm from `ccap2012.csv` (remove NAs). Report both sets of statistics side by side. What do you notice? (1–2 sentences)

```{r}
gay_r_baseline <- gay_r %>%
  filter(study == 1) %>%
  drop_na(therm1)

gay_r_baseline_summary <- gay_r_baseline %>%
  summarize(
    mean= mean(therm1), 
    median = median(therm1), 
    sd = sd(therm1))

ccap_baseline_summary <- ccap %>%
  drop_na(gaytherm) %>%
  summarize(
    mean = mean(gaytherm), 
    median = median(gaytherm), 
    sd = sd(gaytherm))

combined_summary <- tibble(
  stat = c("Mean", "Median", "SD"),
  gay_r = c(
    gay_r_baseline_summary$mean,
    gay_r_baseline_summary$median,
    gay_r_baseline_summary$sd
  ),
  ccap = c(
    ccap_baseline_summary$mean,
    ccap_baseline_summary$median,
    ccap_baseline_summary$sd
  )
)

combined_summary
```

*Looking at the test statistics of both datasets side by side, it seems that the mean, median, and sd between the two are extremely similar in terms of thermometer scoring and have lots of overlap. It would appear that their samples are pretty balanced and similar to one another...*

### Q2.3 (5 points): Create side-by-side histograms comparing the LaCour Study 1 baseline (therm1) and the CCAP data (gaytherm). Use the same bin breaks for both: breaks = seq(0, 100, by = 5).

*Hint*: Combine the two datasets into one data frame with a column indicating the source, then use `ggplot()` with `facet_wrap()`.

Do these look like they came from different populations? (2–3 sentences)

```{r}
# Combine datasets
combined_data <- bind_rows(
  gay_r_baseline %>%
    transmute(therm = therm1, source = "LaCour Study 1"),
  ccap %>%
    drop_na(gaytherm) %>%
    transmute(therm = gaytherm, source = "CCAP"))

# Plot side-by-side histograms
ggplot(combined_data, aes(x = therm)) +
  geom_histogram(breaks = seq(0, 100, by = 5), fill = "lightblue", color = "skyblue") +
  facet_wrap(~ source) +
  labs(
    title = "Baseline Feeling Thermometer: LaCour Study 1 vs CCAP",
    x = "Feeling Thermometer",
    y = "Count") +
  theme_minimal()
```

*The distributions of both datasets look very similar, with the main difference being in the number of observations. Otherwise, the heaping towards the middle value at 50 and the heaping towards the more extreme ends of 0 and 100 track across both distributions, making them look like they came from very similar populations.*

### 2.2 Re-test Reliability

### Q2.4 (6 points): For both the Study 1 and Study 2 control groups (“No Contact”), calculate:

(a) The correlation between `therm1` and `therm2` (use `cor()` with use = "`complete.obs`")
(b) The standard deviation of the within-person change (`therm2` - `therm1`)

Present your results side by side. A re-test correlation of 0.95–0.97 is typical for real feeling thermometer data. Which study shows a more plausible pattern, and why? (3–4 sentences)

```{r}
gay_r_nc_s1_summary <- gay_r %>%
  drop_na(therm1, therm2) %>%
  filter(study == 1 & treatment == "No Contact") %>%
  summarize(
    cor = cor(therm1, therm2, use =  "complete.obs"), 
    sd = sd(therm2 - therm1))
  
gay_r_nc_s2_summary <- gay_r %>%
  drop_na(therm1, therm2) %>%
  filter(study == 2 & treatment == "No Contact") %>%
  summarize(
    cor = cor(therm1, therm2, use =  "complete.obs"), 
    sd = sd(therm2 - therm1))

combined_s1_s2_summary <- tibble(
  stat = c("Correlation", "SD of Within-Person Change"),
  gay_r_control_s1 = c(
    gay_r_nc_s1_summary$cor,
    gay_r_nc_s1_summary$sd
  ),
  gay_r_control_s2 = c(
    gay_r_nc_s2_summary$cor,
    gay_r_nc_s2_summary$sd
  )
)

combined_s1_s2_summary
```

*Study 2 shows a more plausible pattern because it's correlation isn't as high (0.972) as it is for Study 1 (0.998), which in Study 1 seems to be too good to be true. Also, it's standard deviation of the within-person change between `therm1` and `therm2` is larger than it was initially.*

### Q2.5 (5 points): Create two scatter plots side by side:

(a) Study 1 control group: `therm1` (x-axis) vs `therm2` (y-axis), with a 45-degree line
(b) Study 2 control group: `therm1` (x-axis) vs `therm2` (y-axis), with a 45-degree line

Describe the differences. Which looks more like real longitudinal data, and why? (3–4 sentences)

```{r}
gay_r_nc_s1 <- gay_r %>%
  drop_na(therm1, therm2) %>%
  filter(study == 1 & treatment == "No Contact")

gay_r_nc_s2 <- gay_r %>%
  drop_na(therm1, therm2) %>%
  filter(study == 2 & treatment == "No Contact")
  
# Combine datasets
combined_nc_studies <- bind_rows(
  gay_r_nc_s1 %>%
    transmute(therm1 = therm1, therm2 = therm2, source = "Study 1"),
  gay_r_nc_s2 %>%
    transmute(therm1 = therm1, therm2 = therm2, source = "Study 2"))

# Plot side-by-side histograms
ggplot(combined_nc_studies, aes(x = therm1, y = therm2)) +
  geom_point(color = "lightblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "blue", linetype = "solid") +
  facet_wrap(~ source) +
  labs(
    title = "Correlation Between `therm1` and `therm2`",
    x = "Feeling Thermometer",
    y = "Count") +
  theme_minimal()
```

*When looking at the correlation between `therm1` and `therm2`, Study 1 shows nearly perfect positive linear correlation as the data falls barely above and below our 45$\degree$ line. In contrast, Study 2 shows positive linear correlation that is not as strong, as depicted by the jitter in the data above and below the 45$\degree$, and with the data being more spread out or sparse in some regions of the Feeling Thermometer. It is clear that Study 2's scatterplot looks like real longitudinal data in comparison to Study 1 because of the natural noise that seems to exist in Study 2, but in addition to the neatness along the line, the heaping in the middle for Study 1 on the scatterplot looks unnaturally piled along the center in comparison to Study 2.*

### 2.3 Putting It All Together

### Q2.6 (6 points): You’ve now found three irregularities:

1.  The baseline distributions match an existing national survey
2.  The re-test correlations are implausibly high (Study 1)
3.  The within-person changes are implausibly small (Study 1)

In 4–5 sentences, explain why these three findings together suggest the data were fabricated rather than collected in a real experiment. Use specific numbers from your analysis. Why would a fabricator produce data with these properties?

*Taken together, these three patterns strongly suggest the data were fabricated rather than collected from real respondents. First, the baseline distribution in Study 1 closely matches the national CCAP survey, with nearly identical summary statistics, which would be unlikely if an independent sample was taken for LaCour's study, so it is more consistent with copying the already existing dataset. Second, the re-test correlation in Study 1 is almost impossibly high at 0.998, which flags as weird because  real attitude measures over time typically show much lower stability due to measurement error and natural variation. Third, the within-person changes are extremely small, with differences clustered near zero and a very small standard deviation, which is unrealistic in real panel data where individuals normally vary more. A fabricator might generate data this way to make the study look clean and convincing—highly stable attitudes with minimal noise, while also using a known national distribution to ensure the baseline appears realistic.*
